Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.}

### 2022-06-03 analysis/paper/shceof.tex
Render

We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period (Baldwin and Thompson 2009).
We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by Campitelli, Díaz, and Vera (2022).
Seasonal indices of the PSA patterns (PSA1 and PSA2) were calculated, in agreement with Mo and Paegle (2001), as the third and fourth leading EOF of seasonal mean anomalies for 500-hPa geopotential heights at SH.

@@

We now explore the relationship between SAM and the cEOFs motivated by the resemblance between cEOFs regression maps and SAM patterns shown in Section~\ref{regressions}.
We then computed the coefficient of determination between the cEOFs time series and the three SAM indices (SAM, A-SAM and S-SAM) defined by Campitelli, Díaz, and Vera (2022) at each vertical level (Fig.~\ref{fig:sam-eof-vertical}).
The SAM index is statistically significant correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
On the other hand, correlations between SAM and the Real cEOF2 are non-significant.

The relationship between the SAM and cEOF1 in the troposphere is explained entirely by the zonally symmetric component of the SAM as shown by the high correlation with the S-SAM below 100 hPa and the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1.
In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
These correlations are consistent with the regression maps of geopotential height in Figure~\ref{fig:eof1-regr-gh} and their comparison with those obtained for SAM, A-SAM and S-SAM by Campitelli, Díaz, and Vera (2022).

In the case of Imaginary cEOF2, its correlation with the SAM for the troposphere is associated to the asymmetric variability of the SAM.
Indeed, the Imaginary cEOF2 shares up to 92\% variance with the A-SAM and only 12\% at most with the S-SAM (Figure~\ref{fig:sam-eof-vertical}.b2).
Such extremely high correlation between A-SAM and Imaginary cEOF2 suggests that the modes obtained in this work are able to characterise the zonally asymmetric component of the SAM described previously by Campitelli, Díaz, and Vera (2022).

@@

Over South America, the Imaginary cEOF2 has positive correlations with precipitation in South Eastern South America (SESA) and central Chile, and negative correlations in eastern Brazil. This correlation field matches the springtime precipitation signature of ENSO (e.g. Cai et al. 2020) and it is also similar to the precipitation anomalies associated with the A-SAM (Campitelli, Díaz, and Vera 2022).

@@

Our method is similar to Goyal et al. (2022) as they also derive an index with amplitude and phase of wave-like activity in the SH with very similar patterns to our cEOF2.
Both indices are likely very similar (a detailed comparison is out of scope for this paper) but cEOF have the advantage of constructing the indices based on patterns that are exactly in quadrature by construction.

@@

@@

\begin{itemize}
\item
  The Oceanic Niño Index is available via NOAA's Climate Prediction Center: \url{https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt}.
\item
  The Dipole Mode Index is available via Global Climate Observing System Working Group on Surface Pressure: \url{https://psl.noaa.gov/gcos_wgsp/Timeseries/Data/dmi.had.long.data}
\end{itemize}

@@

\leavevmode\vadjust pre{\hypertarget{ref-campitelli2022}{}}%
Campitelli, Elio, Leandro B. Díaz, and Carolina Vera. 2022. {``Assessment of Zonally Symmetric and Asymmetric Components of the {Southern Annular Mode} Using a Novel Approach.''} \emph{Climate Dynamics} 58 (1): 161--78. \url{https://doi.org/10.1007/s00382-021-05896-5}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-era5}{}}%
Hersbach, H., B. Bell, P. Berrisford, G. Biavati, A. Horányi, J. Muñoz Sabater, J. Nicolas, et al. 2019. {``{Era5} Monthly Averaged Data on Pressure Levels from 1979 to Present.''} \emph{{Copernicus Climate Change Service} ({C3s}) {Climate Data Store} ({CDS})} (Accessed on {\(<\)}07-09-2021{\(>\)}). \url{https://doi.org/10.24381/cds.6860a573}.

>>>>

We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period (Baldwin and Thompson 2009).
We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by Campitelli, Díaz, and Vera (2022a).
Seasonal indices of the PSA patterns (PSA1 and PSA2) were calculated, in agreement with Mo and Paegle (2001), as the third and fourth leading EOF of seasonal mean anomalies for 500-hPa geopotential heights at SH.

@@

We now explore the relationship between SAM and the cEOFs motivated by the resemblance between cEOFs regression maps and SAM patterns shown in Section~\ref{regressions}.
We then computed the coefficient of determination between the cEOFs time series and the three SAM indices (SAM, A-SAM and S-SAM) defined by Campitelli, Díaz, and Vera (2022a) at each vertical level (Fig.~\ref{fig:sam-eof-vertical}).
The SAM index is statistically significant correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
On the other hand, correlations between SAM and the Real cEOF2 are non-significant.

The relationship between the SAM and cEOF1 in the troposphere is explained entirely by the zonally symmetric component of the SAM as shown by the high correlation with the S-SAM below 100 hPa and the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1.
In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
These correlations are consistent with the regression maps of geopotential height in Figure~\ref{fig:eof1-regr-gh} and their comparison with those obtained for SAM, A-SAM and S-SAM by Campitelli, Díaz, and Vera (2022a).

In the case of Imaginary cEOF2, its correlation with the SAM for the troposphere is associated to the asymmetric variability of the SAM.
Indeed, the Imaginary cEOF2 shares up to 92\% variance with the A-SAM and only 12\% at most with the S-SAM (Figure~\ref{fig:sam-eof-vertical}.b2).
Such extremely high correlation between A-SAM and Imaginary cEOF2 suggests that the modes obtained in this work are able to characterise the zonally asymmetric component of the SAM described previously by Campitelli, Díaz, and Vera (2022a).

@@

Over South America, the Imaginary cEOF2 has positive correlations with precipitation in South Eastern South America (SESA) and central Chile, and negative correlations in eastern Brazil. This correlation field matches the springtime precipitation signature of ENSO (e.g. Cai et al. 2020) and it is also similar to the precipitation anomalies associated with the A-SAM (Campitelli, Díaz, and Vera 2022a).

@@

The method used in this study have similarities to the one used by Goyal et al. (2022) as they construct an index of amplitude and phase of zonal wave 3-like variability by combining the two leading EOFs of meridional wind anomalies.
The patterns obtained by them bear high resemblance with cEOF2.
Although a detailed comparison is out of scope for this paper, the cEOF analysis has the advantage of constructing the indices based on patterns that are exactly in quadrature by construction.

@@

\hypertarget{availability-of-data-and-material}{%
\subsection*{Availability of data and material}\label{availability-of-data-and-material}}
\addcontentsline{toc}{subsection}{Availability of data and material}

All data used in this paper available in a version-controlled repository (Campitelli, Díaz, and Vera 2022b).
It is also freely available from their respective sources:

\begin{itemize}
\item
  ERA5 data can be obtained via the Copernicus Climate Data Store (\url{https://cds.climate.copernicus.eu/cdsapp/\#!/dataset/reanalysis-era5-pressure-levels-monthly-means/}).
\item
  ERSSTv5 can be obtained via NOAA's NCEI websiste at \url{https://www.ncei.noaa.gov/access/metadata/landing-page/bin/iso?id=gov.noaa.ncdc:C00927}
\item
  CMAP Precipitation data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at \url{https://psl.noaa.gov/data/gridded/data.cmap.html}.
\item
  The Oceanic Niño Index is available via NOAA's Climate Prediction Center: \url{https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt}.
\item
  The Oceanic Niño Index is available via NOAA's Climate Prediction Center: \url{https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt}.
\item
  The Dipole Mode Index is available via Global Climate Observing System Working Group on Surface Pressure: \url{https://psl.noaa.gov/gcos_wgsp/Timeseries/Data/dmi.had.long.data}
\end{itemize}

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-campitelli2022}{}}%
Campitelli, Elio, Leandro B. Díaz, and Carolina Vera. 2022a. {``Assessment of Zonally Symmetric and Asymmetric Components of the {Southern Annular Mode} Using a Novel Approach.''} \emph{Climate Dynamics} 58 (1): 161--78. \url{https://doi.org/10.1007/s00382-021-05896-5}.

\leavevmode\vadjust pre{\hypertarget{ref-campitelli2022a}{}}%
Campitelli, Elio, Leandro Díaz, and Carolina Vera. 2022b. {``Data for "{Revisiting} the {Austral Spring Extratropical Southern Hemisphere} Zonally Asymmetric Circulation Using Complex {Empirical Orthogonal Functions}".''} {Zenodo}. \url{https://doi.org/10.5281/zenodo.6612324}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-era5}{}}%
Hersbach, H., B. Bell, P. Berrisford, G. Biavati, A. Horányi, J. Muñoz Sabater, J. Nicolas, et al. 2019. {``{Era5} Monthly Averaged Data on Pressure Levels from 1979 to Present.''} \emph{{Copernicus Climate Change Service} ({C3s}) {Climate Data Store} ({CDS})} (Accessed on {\(<\)}03-06-2022{\(>\)}). \url{https://doi.org/10.24381/cds.6860a573}.
